Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument

### 2014-10-17 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
draft update

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} imply what is called the modal collapse
\cite{Sobel1987,SobelBook} (formula $\varphi \rightarrow \Box \varphi$
holds for any formula $\varphi$ and not just for $\exists x. God(x)$
as intended). This fact, which has recently been confirmed with
higher-order automated theorem provers \cite{J30,ECAI-2014}, has led
to strong criticism of the argument, and it has therefore stimulated
attempts to remedy the problem. Some authors, including Hajek
\cite{Hajek_der_Mathematiker,Hajek_Magari_and_others}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordahl \cite{}, have proposed  to adapt G\"odel's axioms (in
particular, controversial axiom A1) and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordahl.  More precisely, we have formalized their
emendations within the proof assistant Isabelle/HOL \cite{Isabelle},
and we have then automated (different variations of) their
emendations with the higher-order reasoners Leo-II \cite{}, Satallax
\cite{} and Nitpick \cite{} available in Isabelle/HOL.\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordahl} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced and
employed in our previous work \cite{J30,ECAI-2014}. In this approach
full comprehension is naturally ``built-in'' since the underlying HOL
supports $\lambda$-abstraction. We summarize the main results of our
experiments.

Anderson's emendation (cf. Appendix \ref{apx:Anderson}) has been
verified for different domain conditions. These variations were
motivated by different authors who commented on Anderson's work. Anderson himself \ldots
\begin{itemize}
\item Constant domain semantics \footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}}, that is, the domains of
  objects for all types (individuals, predicates, etc.) stays the same
  in all worlds. We consider constant domain semantics as the standard
  setting.
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_varying_domain.thy}} for the type of indivduals individuals
  (and constant domain for all other types). Different authors have pointed out that 
  Ande
\item The foll
and constant 
has been verified for constant domain\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_const.thy}} and varying
  domain semantics
\item 
\end{itemize}

 that is, the following can be proved by the HOL-ATPs:

G\"odel's and Scott's versions of the proof are not explicit about the
particular comprehension principles they assume.  Some authors,
including Hajek \cite{Hajek_der_Mathematiker,Hajek_Magari_and_others},
have suggested to consider cautious conprehension principles in order
to escape the modal collapse.

>>>>

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} imply what is called the modal collapse
\cite{Sobel1987,SobelBook} (formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. God(x)$ as intended). This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{J30,ECAI-2014}, has led to strong criticism of the argument, and
it has therefore stimulated attempts to remedy the problem. Some
authors, including Hajek
\cite{Hajek_der_Mathematiker,Hajek_Magari_and_others}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordahl \cite{}, have proposed to adapt G\"odel's axioms (in
particular, controversial axiom A1) and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordahl.  More precisely, we have formalized their
emendations within the proof assistant Isabelle/HOL \cite{Isabelle},
and we have then automated (different variations of) their emendations
with the higher-order reasoners Leo-II \cite{}, Satallax \cite{},
Metis \cite{} and Nitpick \cite{}; these systems are available in
Isabelle/HOL.  Our formalizations\footnote{The formalizations are
  available in the subdirectories \url{Anderson} and \url{Bjordahl} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced and employed in our previous
work \cite{J30,ECAI-2014}. In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction.

We summarize the main results of our experiments, and we start with
Anderson's emendation (cf. Appendix \ref{apx:Anderson}), which we have
analysed for different domain conditions. These variations were
motivated by various comments on Anderson's work in the literature. 
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}} (the individual
  domains remain the same in all worlds): G\"odel's theorems C1, and
  T3 can easily be automated (in logics K and KB, respectively) ,
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B, the latter already in K); a trivial countermodel (two
  worlds, one individual; this is simpler than
          the countermodel with two individuals and two worlds
          presented by Anderson, 1990) to MC generated by Nitpick (for all mentioned
  logics); this model also proves consistency.
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_varying_domain.thy}} (the domains for
  indivduals may vary from world to world, all other domains remain
  constant; different authors have pointed out that varying domain
  quantifiers should be used in Anderson's work; here we consistently
  used varying domain for all individual quantifiers; the varying
  domain quantifiers have been modeled by using an explicit
  ``Existence'' relation between indiviudals and worlds): we obtain
  the same results as  above
\item Mixed variant\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_mixed_domain.thy}} (varying domain
  quantifiers are used exclusively in the definitions of essence and
  NE; see the comments by Fitting): G\"odel's axioms A4 and A5 remain
  redundant (the former one in
  logic K4B, the latter already in K).
\end{itemize}
